### 2017-05-07 2017mergingmerges/section/04-Fusion.tex
Fix section on Fusability after the Great Cutting

The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.

Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion. We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process. Each $Label_F$ consists of a $Label$ from one of the source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$. The definition of $Label$ is now recursive. While fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information. Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.

@@

We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with channel @c@. We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
(We used a fresh name back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion.)

Still in Fig.~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes. Type @in2@ indicates that the two processes @pull@ from the same channel, and these actions must then be coordinated. Type @in1@ indicates that only a single process pulls from the channel. Type @in1out1@ indicates that one process pushes to the channel and the other pulls. Type @out1@ indicates that the channel is pushed to by a single process. Each output channel is uniquely owned and cannot be pushed to by more than one process.

@@

\smallskip
% -------------------------------------
Fig.~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes together, constructing a result process that does the job of both. We start with a joint label $l_0$ formed from the initial labels of the two source processes. We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next. The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states. As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no new states can be discovered.

@@

% -------------------------------------
Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.

Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@. This helps collect @jump@ instructions together, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle. As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.

@@

% -------------------------------------
\smallskip
Fig.~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction. This function takes the map of channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.

@@

Still in Fig.~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.

@@

Clause (LocalDrop) applies when the left process wants to drop the current value from that it read from an unshared input channel, which requires no coordination.

@@

\smallskip
Fig.~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states.
Function \ti{swaplabels} flips the order of the compound labels in an instruction.

>>>>

Our core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.

Figure~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion. We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process. Each $Label_F$ consists of a $Label$ from one of the source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$. The definition of $Label$ is now recursive. While fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information. Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.

@@

We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with \mbox{channel @c@}. We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
(We used a fresh name back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion.)

Still in Figure~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes. Type @in2@ indicates that the two processes @pull@ from the same channel, so these actions must be coordinated. Type @in1@ indicates that only a single process pulls from the channel. Type @in1out1@ indicates that one process pushes to the channel and the other pulls. Type @out1@ indicates that the channel is pushed to by a single process. Each output channel is uniquely owned and cannot be pushed to by more than one process.

@@

\smallskip
% -------------------------------------
Figure~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes, constructing a result process that does the job of both. We start with a joint label $l_0$ formed from the initial labels of the two source processes. We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next. The possible destination labels of that instruction (computed with $outlabels$ from Figure~\ref{fig:Fusion:Utils}) define new joint labels and reachable states. As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no new states can be discovered.

@@

% -------------------------------------
Figure~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.

Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@. This helps collect @jump@ instructions together, which are then easier for post-fusion optimization to handle (\S\ref{s:Optimisation}). As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Figure~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.

@@

% -------------------------------------
\smallskip
Figure~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction. This function takes the map of channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.

@@

Still in Figure~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.

@@

Clause (LocalDrop) applies when the left process wants to drop the current value that it read from an unshared input channel, which requires no coordination.

@@

% NOTE BL: commented out to save space. The figure is mentioned earlier.
% \smallskip
% Figure~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned. Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes. Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states. Function \ti{swaplabels} flips the order of the compound labels in an instruction.

% -----------------------------------------------------------------------------
\subsection{Fusibility}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, the final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before pushing values to @s2@. On the other hand, @zip@ needs to pull a \emph{single} value from each of its inputs alternately.

Dynamically, if we were to execute the first fused result process, and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstream @zip@ process the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this manifests as the failure of the $tryStepPair$ function from Figure~\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.

% On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.

% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.

% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.

When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse.
